\section{Exercises}

\begin{exercises}
  \item Define an interface \verb+Asset+ as an abstract class.
        One should be able to query the value of an asset for
        a supplied \verb+Date+. Define the concrete assets,
        \verb+Vehicle+ and \verb+Property+. When creating a 
        \verb+Vehicle+ one should specify the purchase price
        and date as well as the write-off period in years.
        Assume the vehicle is written off linearly over the 
        write-off period. Properties also have a purchase 
        price and date. Assume, for sake of simplicity, that
        properties neither loose nor gain value. Define 
        further a portfolio class to which you can add assets.
        The portfolio itself is, also an asset. Now create two 
        portfolios, each with a property and a vehicle and add
        the one portfolio to the other (after all, a portfolio
        is an asset too). Query todays value of the portfolio.
        To help you work with dates look at the following 
        functions defined in \verb+time.h+:
        \begin{description}
          \item[difftime] which calculates the time in milliseconds
                          between two times.
          \item[mktime] which makes a \verb+time_t+ from a \verb+tm+
                        structure.
          \item[localtime] which creates a \verb+time_t+ representing 
                           the current date/time.
        \end{description}
        For simplicity, assume each year has 365 days.
\end{exercises}
